---
title: "Homework 6"
author: "Lauren Richter"
date: 2021-12-04
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, collapse = TRUE)

library(tidyverse)
```

## Problem 1

In this problem, you will analyze data gathered to understand the effects of several variables on a child’s birthweight. This dataset, available here, consists of roughly 4000 children and includes the following variables:

- `babysex`: baby’s sex (male = 1, female = 2)
- `bhead`: baby’s head circumference at birth (centimeters)
- `blength`: baby’s length at birth (centimeteres)
- `bwt`: baby’s birth weight (grams)
- `delwt`: mother’s weight at delivery (pounds)
- `fincome`: family monthly income (in hundreds, rounded)
- `frace`: father’s race (1 = White, 2 = Black, 3 = Asian, 4 = Puerto Rican, 8 = Other, 9 = Unknown)
- `gaweeks`: gestational age in weeks
- `malform`: presence of malformations that could affect weight (0 = absent, 1 = present)
- `menarche`: mother’s age at menarche (years)
- `mheight`: mother’s height (inches)
- `momage`: mother’s age at delivery (years)
- `mrace`: mother’s race (1 = White, 2 = Black, 3 = Asian, 4 = Puerto Rican, 8 = Other)
- `parity`: number of live births prior to this pregnancy
- `pnumlbw`: previous number of low birth weight babies
- `pnumsga`: number of prior small for gestational age babies
- `ppbmi`: mother’s pre-pregnancy BMI
- `ppwt`: mother’s pre-pregnancy weight (pounds)
- `smoken`: average number of cigarettes smoked per day during pregnancy
- `wtgain`: mother’s weight gain during pregnancy (pounds)


```{r load_data, collapse=TRUE}
bw_data = read_csv("data/birthweight.csv") %>%
  janitor::clean_names("snake") %>%
  mutate(babysex = as_factor(babysex),
         frace = as_factor(frace), 
         mrace = as_factor(mrace))

# check for the number of missing variables using sum(is.na(bw_data))
```

Data tidying involved name cleaning, variable conversion, and checking for missing data.  Variable names were cleaned to ensure they were consistent (using snake case, although because of the initial formatting of variables, snake case resulted only in lowercase variable names sans underscores). The variables `babysex`,`frace`, and `mrace` were converted from numeric to factor variables. There were `r sum(is.na(bw_data))` missing variables in the data set.  

A regression model for birthweight

```{r bw_model}

# value = sex, head, length, GA, malform, pnumlbw pnumsga smoken wtgain

```


Propose a regression model for birthweight. This model may be based on a hypothesized structure for the factors that underly birthweight, on a data-driven model-building process, or a combination of the two. Describe your modeling process and show a plot of model residuals against fitted values – use add_predictions and add_residuals in making this plot.

Compare your model to two others:

One using length at birth and gestational age as predictors (main effects only)

```{r length_v_ga}
# value = B*length + C*GA
```

One using head circumference, length, sex, and all interactions (including the three-way interaction) between these

```{r hc_length_sex}
# value = B*hc + C*length + D*sex + E*hc*length*sex + F*hc*length + G*hc*sex + H*length*sex
```
Make this comparison in terms of the cross-validated prediction error; use crossv_mc and functions in purrr as appropriate.

Note that although we expect your model to be reasonable, model building itself is not a main idea of the course and we don’t necessarily expect your model to be “optimal”.

